Effects of the body force on the evacuation dynamics

### 2020-02-03 paper.tex
commit para hacer pull

The Social Force Model (SFM) models the crowd dynamics 
considering three kinds of force: desired forces, social forces, and physical forces.
In its original version, the SFM, addresses two physical forces as essential: 
 the ``body force'' and the ``sliding friction''. Both are 
inspired by granular interactions and were claimed to be necessary  
for attaining the particular effects in panicking crowds \cite{helbing_2000}. 
The ``sliding friction'' actually proved to be an essential feature of the 
``faster-is-slower'' effect, although the role of the ``body force'' appears, 
at a first instance, not so clear \cite{dorso_2005,dorso_2007,dorso_2011}. \\

The existence of a ``body force'' in the context of highly dense crowds (say, 5 
to 10$\,$people/m$^2$) is a commonsense matter \cite{henein_2007,fruin_1993}. 
Researchers, however, question the numerical setting for this force in 
the SFM context \cite{lakoba_2005}. As a matter of fact, the usual setting by 
Helbing prevents the excessive overlapping among pedestrians, but it is known to 
accomplish artificially high force levels 
\cite{helbing_2000,lakoba_2005,langston_2006,lin_2017}. The force estimates 
from the SFM appear to be remarkably higher with respect to the reported real 
life data (say, an order of magnitude). The crowd motion simulations, however, 
present quite realistic results \cite{lakoba_2005,langston_2006,dorso_2017}. 
The 
point seems to be that the SFM focuses on the ``collective behavior'' due 
to clogging, missing the ``individualistic'' perspective  of single pedestrians 
or very small groups \cite{helbing_2000,henein_2007,narain_2009}.  \\

@@

A unique (and universal) setting appears not available yet to
our knowledge. 
 The reason is that different numerical settings can lead to 
the same crowd dynamics. Actually, only a small set of dimensionless 
``numbers'' control the crowd dynamics \cite{dorso_2019}. These are similar to 
those encountered in other active matter systems (P\'eclet number, etc.)  
\cite{marchetti_2014}. We may hypothesize that while the dimensionless 
``numbers'' provide some kind of control on the 
collective behavior in 
crowds, only a few numerical settings can attain an ``individualistic'' 
meaning. 
 \\

@@

We now widen the investigation on the parameter settings to include the one 
associated to the body force. We will  explore 
 the complex interplay between 
the body force and the sliding friction among pedestrians. Recall that the 
interplay dynamics is not directly controlled by the numerical setting, but 
through dimensionless ``numbers'', where the model parameters appear mixed 
between each other. Thus, this step up offers a challenge to the 
``individualistic'' meaning of the parameter's setting. \\

The investigation is organized as follows. We first recall the available 
experimental values on the body force and the sliding friction (see Section 
\ref{experimental}). Secondly, we introduce the reduced-in-units SFM and the 
three dimensionless numbers that control the crowd dynamics (see Section 
\ref{background}). We present our numerical simulations in Section 
\ref{results}. For the sake of clarity, this Section is 
separated into two major parts: the bottleneck scenario and the corridor 
scenario in Sections \ref{bottleneck} and \ref{corridor}, respectively. 
Section \ref{conclusions} opens a detailed discussion from results in Section \ref{results}
and resumes our main conclusions.

@@

The complex behavior of pedestrians features
 either his (her) feelings and 
the environmental conditions. The former is expressed, for example, by his 
(her) moving ``attitude'' (say, self-assuredness). The latter brings out the 
observed separation between pedestrians. Also, the ``contacts'' between 
individuals address some kind of ``unwanted'' slowing down. All these 
observed patterns are commonly quantified in the literature into a set of 
characteristic  parameters. The experimental meaning of these parameters is as 
follows.

\begin{itemize}
\item[(i)] The walking attitude of a pedestrian may appear somewhat 
  ``aggressive'' if he (she) reacts actively to unexpected behaviors 
\cite{lakoba_2005,helbing_1995}. The smaller the reaction time, the more 
 aggressive observed posture. The associated parameter to this 
behavior is the relaxation or characteristic time $\tau$ 
\cite{johansson_2009,helbing_2000}.

\item[(ii)] Despite the reactive attitude $\tau$, the pedestrian addresses a 
``free'' (undisturbed) walking speed $v_d$. This speed expresses his (her) 
motivation or intention to reach a certain destination (as comfortable as 
possible). Observations commonly associate $0.6\,$m/s, $1\,$m/s or $1.5\,$m/s 
to relaxed, normal or nervous walking speeds, respectively  
\cite{helbing_1995,helbing_2000,li_2015}.

@@

Table~\ref{table_data} shows a few empirical values
 for the most common 
parameters. More data is available throughout the literature (see, for example, 
Refs.~\cite{hoogendoorn_2007,seyfried_2007,johansson_2007,moussaid_2009,
luber_2010,seer_2014,li_2015} ). We intentionally omitted data that assumes a 
specific mathematical model. The exhibited values should also be considered as a 
general purpose approach, since no distinction has been made on age, gender or 
cultural habits. \\

@@

\hline
\end{tabular}
\caption{The experimental data for the pedestrian parameters, as explained in 
Section~\ref{experimental}. The magnitude $v$ means the actual pedestrian velocity 
(m/s). The magnitude $x$ means the compression length (m). The upper row for 
Ref.~\cite{song_2019} corresponds to data acquired in winter and the lower row to 
data acquired in summer. The asterisk ($^*$) corresponds to reasonable 
estimates from the authors. }
\label{table_data}
\end{table}

A first examination of the figures in Table~\ref{table_data} shows that the 
choice $\tau\simeq0.6\,$s seems to be a reasonable estimation for the 
relaxation time, although this may vary with respect to gender or culture 
\cite{siddharth_2018}. Additionally, we confirm that normal pedestrians attain 
desired velocities around $1.3\,$m/s. \\

The reports from Refs.~\cite{seyfried_2007,weidmann_1992} do not include any 
values for the compressibility $k_n$ since these experiments were carried out 
under low density conditions.  The minimum (perceived) step distance 
is $0.36\,$m according to Ref.~\cite{seyfried_2007}, but the
 pedestrians seem to require larger 
distances when they walk faster. The commonly accepted value $B\simeq 
0.5\,$m is somewhat valid for walking speeds under $0.5\,$m/s 
\cite{seyfried_2007}. Higher walking speeds (say, $1\,$m/s) will require 
a step distance of $1.3\,$m for the pedestrians to feel that there is enough 
space to move along.     \\

The reported data from Ref.~\cite{song_2019} correspond to the crowded 
environment of the Beijing subway. This environment was not suitable for 
providing information on the step distance $B$, but estimates for the 
desired speed and the body compressibility could be achieved. The reported 
magnitude $k_n$ assesses either the clothes and the body compressibility. The 
final value, though, is linearly related to the compression $x$. \\

We measured the maximum attainable forces during the spring of 2019 at the subway in Buenos Aires, 
Argentina. Our preliminary results show that pedestrians feel ``uncomfortable'' 
whenever a body force ranging from $5$ to $20\,$N is applied for at least ten 
minutes. Short lasting forces (say, less than 4 minutes) may also be perceived 
as ``uncomfortable'' for values ranging from $10$ to $30\,$N. We also recorded body 
forces up to $60\,$N during very short ``hits''. The comparison with the 
fittings provided by Ref.~\cite{song_2019} shows that these magnitudes 
accomplish densities around 5 people/m$^2$.     \\

The maximum (realistic) compressions may be computed from the hook relation 
$F(x)=k_n(x)\,x$ and the compressibility $k_n(x)$ reported in Table~\ref{table_data}. 
An ``uncomfortable'' body force  $10\,$N $-$ $30\,$N can address compressions in the 
range of $0.030-0.055\,$m. Also, a ``hitting'' force of $60\,$N can address 
compressions between $0.045$ and $0.065\,$m. Thus, according to 
Table~\ref{table_data}, we may expect experimental values for $k_n(x)$ up to 
$1000\,$kg/s$^2$ (for $F=30\,$N), or, up to $1400\,$kg/s$^2$ (for $F=60\,$N). \\

Besides, no reliable values for the sliding friction $\kappa$ appears to be 
available in the literature (to our knowledge). We may presume, however, that 
the sliding friction approaches a fraction of the body force. We will come back 
to this issue in Section~\ref{background}. \\

@@

\noindent where the subscript $j$ corrresponds to any neighboring pedestrian or 
the walls. The three forces $\mathbf{f}_d$, $\mathbf{f}_s$ and $\mathbf{f}_g$ 
are different in nature. The desire force $\mathbf{f}_d$ represents the 
acceleration (or deceleration) of the pedestrian due to his (her) own willings. 
The social force $\mathbf{f}_s$, instead, describes the tendency of the 
pedestrians to stay away from each other. The granular force $\mathbf{f}_g$ 
stands for either the sliding friction and the compression between 
pedestrians. \\

@@

The net distance (overlap) $|R_{ij}-r_{ij}|$ scales to the parameter $B$ in the 
expression (\ref{eqn_social}). This parameter plays the role of a fall-off 
length within the model, and thus, it may be somewhat connected to the 
(perceived) step distance mentioned in Section~\ref{experimental}. The 
parameter $A$, however, does not provide any direct link to other parameters 
mentioned there. \\

@@

The numerical setting of the parameters may affect the dynamics of the 
pedestrians. Some settings, however, yield similar collective dynamics.
We introduce unit-less magnitudes 
and proceed straightforward as indicated in \ref{appendix1}. We realize from 
\ref{appendix1} that only three (unit-less) parameters  are the true 
``control'' parameters for the collective dynamics. These are $\mathcal{A}$, 
$\mathcal{K}$, $\mathcal{K}_c$ as defined in \ref{appendix1}. Recall that 
$\mathcal{A}$ and $\mathcal{K}$ are precisely the same as in 
Ref.~\cite{dorso_2019}, but a novel $\mathcal{K}_c$ parameter has been 
introduced due to the body force.   \\

@@

\section{\label{hypotheses}Working hypotheses and procedure}

\begin{itemize}

\item 1)  Nos enfocamos en estudiar principalmente los efectos del parametro kn. Variamos kn en la mayoria de los resultados, salvo en el diagrama fundamental donde tambien variamos kt.

\item 2)  No usamos anisotropia. Estamos interesados en las situaciones de alta densidad donde suponemos que la anisotropia no juega ningun papel relevante.

\item 3) Las simulaciones numéricas las hacemos con LAMMPS usando módulos creados por nosotros para calcular las fuerzas del modelo de Helbing.

\item 4) Exploramos situaciones de muy alta densidad (muy alta presion), sabemos que en estas situaciones pueden producirse caidas y desmayos pero en este paper ignoramos esos efectos.

>>>>

\textcolor{red}{The Social Force Model (SFM) is a pedestrian dynamics model that
considers three kinds of forces: desired forces, social forces, and physical forces.
In its original version, the SFM, addresses two physical forces as essential: 
\sout{The Social Force Model (SFM) addresses two physical forces as 
\textit{essential}:}} the ``body force'' and the ``sliding friction''. Both are 
inspired by granular interactions and were claimed to be necessary  
for attaining the particular effects in panicking crowds \cite{helbing_2000}. 
The ``sliding friction'' actually proved to be an essential feature of the 
``faster-is-slower'' effect, although the role of the ``body force'' appears, 
at a first instance, not so clear \cite{dorso_2005,dorso_2007,dorso_2011}. \\

The existence of a ``body force'' in the context of highly dense crowds (say, 5 
to 10$\,$people/m$^2$) is a commonsense matter \cite{henein_2007,fruin_1993}. 
Researchers, however, question the numerical setting for this force in 
the SFM context \cite{lakoba_2005}. As a matter of fact, the usual setting by 
Helbing prevents the \textcolor{red}{excessive} overlapping among pedestrians, but it is known to 
accomplish artificially high force levels 
\cite{helbing_2000,lakoba_2005,langston_2006,lin_2017}. The force estimates 
from the SFM appear to be remarkably higher with respect to the reported real 
life data (say, an order of magnitude). The crowd motion simulations, however, 
present quite realistic results \cite{lakoba_2005,langston_2006,dorso_2017}. 
The 
point seems to be that the SFM focuses on the \textcolor{red}{``collective behavior''
 \sout{``incoordination phenomenon''}} due 
to clogging, missing the ``individualistic'' perspective  of single pedestrians 
or very small groups \cite{helbing_2000,henein_2007,narain_2009}.  \\

@@

\textcolor{red}{A unique (and universal) setting appears not available yet. 
\sout{A meaningful (individualistic) numerical setting appears not available yet (to 
our knowledge).}} The reason is that different numerical settings can lead to 
the same crowd dynamics. Actually, only a small set of dimensionless 
``numbers'' control the crowd dynamics \cite{dorso_2019}. These are similar to 
those encountered in other active matter systems (P\'eclet number, etc.)  
\cite{marchetti_2014}. We may hypothesize that while the dimensionless 
``numbers'' provide some kind of control on the \textcolor{red}{\sout{
``incoordination phenomenon''} collective behavior} in 
crowds, only a few numerical settings can attain an ``individualistic'' 
meaning. 
 \\

@@

We now widen the investigation on the parameter settings to include the one 
associated to the body force. We will \textcolor{red}{\sout{focus} explore} 
on the complex interplay between 
the body force and the sliding friction among pedestrians. Recall that the 
interplay dynamics is not directly controlled by the numerical setting, but 
through dimensionless ``numbers'', where the model parameters appear mixed 
between each other. Thus, this step up offers a challenge to the 
``individualistic'' meaning of the parameter's setting. \\

The investigation is organized as follows. We first recall the available 
experimental values on the body force and the sliding friction (see Section 
\ref{experimental}). Secondly, we introduce the reduced-in-units SFM and the 
three dimensionless numbers that control the crowd dynamics (see Section 
\ref{background}). We present our numerical simulations in Section 
\ref{results}. For the sake of clarity, this Section is 
separated into two major parts: the bottleneck scenario and the corridor 
scenario in Sections \ref{bottleneck} and \ref{corridor}, respectively. 
Section \ref{discussion} opens a detailed discussion raising from of the results 
in Section \ref{results}. The last Section closes the investigation with our 
main conclusions.

@@

The complex behavior of pedestrians \textcolor{red}{\sout{feature} features}
 either his (her) feelings and 
the environmental conditions. The former is expressed, for example, by his 
(her) moving ``attitude'' (say, self-assuredness). The latter brings out the 
observed separation between pedestrians. Also, the ``contacts'' between 
individuals address some kind of ``unwanted'' slowing down. All these 
observed patterns are commonly quantified in the literature into a set of 
characteristic  parameters. The experimental meaning of these parameters is as 
follows.

\begin{itemize}
\item[(i)] The walking attitude of a pedestrian may appear somewhat 
 \textcolor{red}{\sout{``assertive''} ``aggressive''} if he (she) reacts actively to unexpected behaviors 
\cite{lakoba_2005,helbing_1995}. The smaller the reaction time, the more 
\textcolor{red}{\sout{assertive or}}  aggressive observed posture. The associated parameter to this 
behavior is the relaxation or characteristic time $\tau$ 
\cite{johansson_2009,helbing_2000}.

\item[(ii)] Despite the reactive attitude $\tau$, the pedestrian addresses a 
``free'' (undisturbed) walking speed $v_0$. This speed expresses his (her) 
motivation or intention to reach a certain destination (as comfortable as 
possible). Observations commonly associate $0.6\,$m/s, $1\,$m/s or $1.5\,$m/s 
to relaxed, normal or nervous walking speeds, respectively  
\cite{helbing_1995,helbing_2000,li_2015}.

@@

Table~\ref{table_data} shows a few empirical \textcolor{red}{\sout{figures} values}
 for the most common 
parameters. More data is available throughout the literature (see, for example, 
Refs.~\cite{hoogendoorn_2007,seyfried_2007,johansson_2007,moussaid_2009,
luber_2010,seer_2014,li_2015} ). We intentionally omitted data that assumes a 
specific mathematical model. The exhibited values should also be considered as a 
general purpose approach, since no distinction has been made on age, gender or 
cultural habits. \\

@@

\hline
\end{tabular}
\caption{The experimental data for the pedestrian parameters, as explained in 
Section~\ref{experimental}. The magnitude $v$ means de pedestrian velocity 
(m/s). The magnitude $x$ means the compression length (m). The upper row for 
Ref.~\cite{song_2019} corresponds data acquired in winter and the lower row to 
datas acquiquired in summer. The asterisk ($^*$) corresponds to reasonable 
estimates from the authors. }
\label{table_data}
\end{table}

A first examination of the figures in Table~\ref{table_data} shows that the 
choise $\tau\simeq0.6\,$s seems to be a reasonable estimation for the 
relaxation time, although this may vary with respect to gender or culture 
\cite{siddharth_2018}. Additionally, we confirm that normal pedestrians attain 
desired velocities around $1.3\,$m/s. \\

The reports from Refs.~\cite{seyfried_2007,weidmann_1992} do not include any 
values for the compressibility $k_n$ since these experiments were carried out 
under low density conditions. \textcolor{red}{\sout{Notice that the} The} 
minimum (perceived) step distance 
is $0.36\,$m \textcolor{red}{according to Ref.}\cite{seyfried_2007}, but the
 pedestrians seem to require larger 
distances when they walk faster. The commonly accepted value $B\simeq 
0.5\,$m is somewhat valid for walking speeds under $0.5\,$m/s 
\cite{seyfried_2007}. Higher walking speeds (say, $1\,$m/s) will require 
a step distance of $1.3\,$m for the pedestrians to feel that there is enough 
space to move along.     \\

The reported data from Ref.~\cite{song_2019} correspond to the crowded 
environment of the Beijing subway. This environment was not suitable for 
providing information on the step distance $B$, but estimates for the 
desired speed and the body compresibility could be achieved. The reported 
magnitude $k_n$ assesses either the clothes and the body compressibility. The 
final value, though, is linearly related to the compression $x$. \\

We measured the maximum attainable forces at the subway in Buenos Aires, 
Argentina. Our preliminary results show that pedestrians feel ``uncomfortable'' 
whenever a body force ranging from $5$ to $20\,$N is applied for at least ten 
minutes. Short lasting forces (say, less than 4 minutes) may also be perceived 
as ``uncomfortable'' for values ranging from $10$ to $30\,$N. We also recorded body 
forces up to $60\,$N during very short ``hits''. The comparison with the 
fittings provided by Ref.~\cite{song_2019} shows that these magnitudes 
accomplish densities around 5 people/m$^2$.     \\

The maximum (realistic) compressions may be computed from the relation 
$F=k(x)\,x$ and the compressibility $k_n(x)$ reported in Table~\ref{table_data}. 
An ``uncomfortable'' body force \textcolor{red}{\sout{of $30\,$N}} can address compressions in the 
range of $0.030-0.055\,$m. Also, a ``hitting'' force of $60\,$N can address 
compressions between $0.045$ and $0.065\,$m. Thus, according to 
Table~\ref{table_data}, \textcolor{blue}{we may expect experimental values for $k_n(x)$ up to 
$1000\,$kg/s$^2$ (for $F=30\,$N), or, up to $1400\,$kg/s$^2$ (for $F=60\,$N).} \\

Besides, no realiable values for the sliding friction $\kappa$ appears to be 
available in the literature (to our knowledge). We may presume, however, that 
the sliding friction approaches a fraction of the body force. We will come back 
to this issue in Section~\ref{background}. \\

@@

\noindent where the subscript $j$ corrresponds to any neighboring pedestrian or 
the walls. The three forces $\mathbf{f}_d$, $\mathbf{f}_s$ and $\mathbf{f}_g$ 
are different in nature. The desire force $\mathbf{f}_d$ represents the 
acceleration (or deceleration) of the pedestrian due to his (her) own willings. 
The social force $\mathbf{f}_s$, \textcolor{red}{\sout{intead} instead}, describes the tendency of the 
pedestrians to stay away from each other. The granular force $\mathbf{f}_g$ 
stands for either the sliding friction and the compression between 
pedestrians. \\

@@

The net distance \textcolor{red}{(overlap)} $|R_{ij}-r_{ij}|$ scales to the parameter $B$ in the 
expression (\ref{eqn_social}). This parameter plays the role of a fall-off 
length within the model, and thus, it may be somewhat connected to the 
(perceived) step distance mentioned in Section~\ref{experimental}. The 
parameter $A$, however, does not provide any direct link to other parameters 
mentioned there. \\

@@

The numerical setting of the parameters may affect the dynamics of the 
pedestrians. Some settings, however, yield similar collective dynamics.
\textcolor{red}{\sout{In order to investigate these similarities, we} We}
 introduce unit-less magnitudes 
and proceed straightforward as indicated in \ref{appendix1}. We realize from 
\ref{appendix1} that only three (unit-less) parameters  are the true 
``control'' parameters for the collective dynamics. These are $\mathcal{A}$, 
$\mathcal{K}$, $\mathcal{K}_c$ as defined in \ref{appendix1}. Recall that 
$\mathcal{A}$ and $\mathcal{K}$ are precisely the same as in 
Ref.~\cite{dorso_2019}, but a novel $\mathcal{K}_c$ parameter has been 
introduced due to the body force.   \\

@@
